The Perils of Exploration under Competition: \\ A Computational Modeling Approach

### 2019-04-30 paper/competing_bandits.tex
Allowed color-edits.
some edits

\documentclass[format=acmsmall, review=true]{acmart}

@@

\newcommand{\HMR}{\term{HMR}}
\newcommand{\HM}{\term{HM}}
% a very useful package for edits and comments, from David Kempe (USC)
%\usepackage{color-edits}
\usepackage[suppress]{color-edits}  % use this to suppress the package
\addauthor{as}{red}    % as for Alex
\addauthor{ga}{blue}  % ga for Guy
\addauthor{sw}{orange} % sw for Steven
% e.g. for Alex, provides \asedit{}, \ascomment{} and \asdelete{}.
\setlength{\tabcolsep}{8pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5} % Default value: 1

@@

\begin{document}
\title[Competing Bandits: The Perils of Exploration under Competition]
{Competing Bandits: \\ The Perils of Exploration under Competition}

\author{Submission 316}

@@

\newpage

>>>>

\documentclass[format=acmsmall, review=false]{acmart}

@@

\newcommand{\HMR}{\term{HMR}}
\newcommand{\HM}{\term{HM}}
% a very useful package for edits and comments, from David Kempe (USC)
\usepackage{color-edits}
%\usepackage[suppress]{color-edits}  % use this to suppress the package
\addauthor{as}{red}    % as for Alex
\addauthor{ga}{blue}  % ga for Guy
\addauthor{sw}{orange} % sw for Steven
% e.g. for Alex, provides \asedit{}, \ascomment{} and \asdelete{}.
\setlength{\tabcolsep}{8pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5} % Default value: 1

@@

\begin{document}
\title[The Perils of Exploration under Competition: A Computational Modeling Approach]
{The Perils of Exploration under Competition: \\ A Computational Modeling Approach}

 \author{Guy Aridor}
 \affiliation{ \institution{Columbia University}}
 \author{Kevin Liu}
 \affiliation{\institution{Columbia University}}
 \author{Aleksandrs Slivkins}
 \affiliation{\institution{Microsoft Research, New York City}}
 \author{Zhiwei Steven Wu}
 \affiliation{\institution{University of Minnesota - Twin Cities}}

 \authorsaddresses{%
  Authors' addresses: Guy Aridor, Columbia University, New York, NY, g.aridor@columbia.edu; Kevin Liu, Columbia University, New York, NY, kevin.liu@columbia.edu; Aleksandrs Slivkins, Microsoft Research, New York, NY, slivkins@microsoft.com; Zhiwei Steven Wu, University of Minnesota - Twin Cities, Minneapolis, MN, zsw@umn.edu}

@@

### 2019-04-30 paper/content/introduction.tex
some edits

\OMIT{\gaedit{Bing experimenting with a ranking algorithm that ends up with a poor user experience could lead to negative short-run consequences in consumers' perceived quality of Bing and lead to more consumers choosing Google over Bing. However, this experimentation can help Bing eventually find a better ranking algorithm that would lead to positive long-run consequences in consumers' perceived quality of Bing and lead to more consumers choosing Bing over Google.}}

@@

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a first-mover. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``incumbent" (the first-mover in temporary monopoly) to permanent duopoly to ``entrant" (late-arriver in temporary monopoly).%
\footnote{\asedit{We consider the ``permanent monopoly" scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}}

@@

%and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.\swcomment{why ``even''?}

@@

\OMIT{ %%%%%%
\swedit{Beyond the welfare implications, we also investigate the barrier to entry that follows from the incumbent's data advantage.  We run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage from the incumbent's ``small head start'' can lead to a significant incumbency advantage, especially when the incumbent commits to a more advanced bandit algorithm. Our results show that even a small degree of data asymmetry can be amplified to be a larger market advantage through competition dynamics. \gadelete{This is in contrast to the data advantage in a static environment such as discussed in \cite{varian2018artificial,bajari2018impact}, in which a small amount of additional data can only provide limited improvement in the predictive accuracy.} The effect is stronger when the incumbent commits to a more advanced bandit algorithm due to
}} %%%%%%
% in a competitive environment the relative differences in the data
% possessed by the firms may lead to a large competitive benefit due
% to the feedback loops discussed previously. In our model this effect
% is amplified when the incumbent commits to a more advanced bandit
% algorithm since, in the incumbency period, the firm gathers data on
% possible actions more efficiently leading to the relative gap upon
% entry to be larger. This is important when exploration can have
% larger reputational consequences.

\OMIT{Our findings on temporary monopoly shed light on the ``first-mover advantage" phenomenon in the digital economy. Being first in the market gives firms free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We investigate which of the two is a stronger barrier to entry. We find that both are strong barriers on their own: removing either one still gives the incumbent a substantially larger share of the market compared to the case where both are removed. However, the data advantage tends to be a stronger barrier when the incumbent commits to a more advanced bandit algorithm. One take-away is that the data advantage is not just about data quantity but also data quality.}

@@

We also consider an alternative choice rule with explicit noise/randomness: a small fraction of users choose a firm uniformly at random.%
\footnote{\asedit{Reputation scores already introduce some noise into users' choices. However, the amount of noise due to this channel is typically small, both in our simulations and in practice, because reputation signals average over many datapoints.}}

 We confirm the theoretical intuition that better algorithms prevail if the parameter -- essentially, the expected number ``random users" -- is sufficiently large. However, we find that this effect is negligible for some smaller but still ``relevant" parameter values.

@@

\asedit{While amenable to simulations, our model appears difficult to analyze. This is for several reasons:
%
intricate feedback loop from performance to reputations to users to performance;
%
mean reputation, most connected to our intuition, is sometimes bad predictor in competition (see Sections~\ref{sec:isolation} and~\ref{sec:revisited});
%
mathematical tools from regret-minimization would only produce ``asymptotic" results, which does not seem to suffice.
We believe that resolving first-order theoretical questions would not add much value to this paper, because of \cite{CompetingBandits-itcs18}. We are in realm of ``asymptotic" theorems about stylized Economic models, which provide some mathematical intuition but no direct bearing on the world. But \cite{CompetingBandits-itcs18} already has an elaborate set of such results, with similar high-level conclusions. So, one would need to go beyond the first-order conclusions to add value.}

@@

The study of competition vs. exploration has been initiated in \cite{CompetingBandits-itcs18}, as discussed above.
\OMIT{Their model differs from ours in two key respects. First, users do not see any signal about firms' past performance, and instead choose between firms according to the Bayesian-expected reward. Second, they vary the strength of competition using assumptions about (ir)rational consumer behavior, whereas we use early entry. Their results are purely theoretical; their model is amenable to proofs but not to numerical experiments. Their high-level conclusion is an inverted-U relationship between competition and innovation that is similar to ours.}
Our setting is also closely related to the ``dueling algorithms" framework \cite{DuelingAlgs-stoc11}, but this framework considers offline / full feedback scenarios whereas we focus on online machine learning problems.

>>>>

@@

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a first-mover. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``incumbent" (the first-mover in temporary monopoly) to permanent duopoly to ``entrant" (late-arriver in temporary monopoly).%
\footnote{\asedit{We consider the ``permanent monopoly" scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}}\footnote{\gaedit{The most common measures of market ``competitiveness" such as the Lerner Index or the Herfindahl-Hirschman Index of a market rely on ex-post observable attributes of a market such as prices or market shares \cite{tirole1988theory}. However, since we have no prices and market shares are an endogenous element of our model, we simply define ``competitiveness" in terms of the number of firms and timing of entry into the market.}}

@@

@@

@@

We also consider an alternative choice rule with explicit noise/randomness: a small fraction of users choose a firm uniformly at random.%
\footnote{\asedit{Reputation scores already introduce some noise into users' choices. However, the amount of noise due to this channel is typically small, both in our simulations and in practice, because reputation signals average over many datapoints.}} We confirm the theoretical intuition that better algorithms prevail if the \gaedit{expected number of ``random" users is sufficiently large.}\gadelete{parameter -- essentially, the expected number ``random users" --is sufficiently large.} However, we find that this effect is negligible for some smaller but still ``relevant" parameter values.

@@

\asedit{While amenable to simulations, our model appears difficult to analyze. This is for several reasons:
%
intricate feedback loop from performance to reputations to users to performance;
%
mean reputation, most connected to our intuition, is sometimes \gaedit{a} bad predictor in competition (see Sections~\ref{sec:isolation} and~\ref{sec:revisited});
%
mathematical tools from regret-minimization would only produce ``asymptotic" results, which do not seem to suffice. We believe that resolving first-order theoretical questions would not add much value to this paper, because of \cite{CompetingBandits-itcs18}.\gadelete{We are in \gaedit{the} realm of ``asymptotic" theorems about stylized economic models, which provide some mathematical intuition but no direct bearing on the world. But \cite{CompetingBandits-itcs18} already has an elaborate set of such results, with similar high-level conclusions. So, one would need to go beyond the first-order conclusions to add value.}}

@@

The study of competition vs. exploration has been initiated in \cite{CompetingBandits-itcs18}, as discussed above.
Our setting is also closely related to the ``dueling algorithms" framework \cite{DuelingAlgs-stoc11}, but this framework considers offline / full feedback scenarios whereas we focus on online machine learning problems.

### 2019-04-30 paper/content/inverted_u.tex
Heavy-tail -> Heavy-Tail
some edits

We investigate the firms' market shares when they choose different algorithms (otherwise, by symmetry both firms get half of the agents). We report the market shares for Heavy-Tail and Needle-in-Haystack instances in Table \ref{sim_table}  (see the first line in each cell), for a range of values of the warm start $T_0$. Table~\ref{tab:duopoly_unif} reports similarly on the Uniform instance. We find that $\DG$ is a weakly dominant strategy for the Heavy-tail and Uniform instances, as long as $T_0$ is sufficiently small. However, \TS is a weakly dominant strategy for the Needle-in-Haystack instance. We find that for a sufficiently small $T_0$, \DG yields more than half the market against \TS,  but achieves similar market share vs. \DG and \DEG. By our tie-breaking rule, (\DG,\DG) is the only pure-strategy equilibrium.

@@

\begin{finding}
\textit{When all firms deploy \DG, and start at the same time, welfare is weakly decreasing as the number of firms decreases.}
\end{finding}

>>>>

We investigate the firms' market shares when they choose different algorithms (otherwise, by symmetry both firms get half of the agents). We report the market shares for Heavy-Tail and Needle-in-Haystack instances in Table \ref{sim_table}  (see the first line in each cell), for a range of values of the warm start $T_0$. Table~\ref{tab:duopoly_unif} reports similarly on the Uniform instance. We find that $\DG$ is a weakly dominant strategy for the Heavy-Tail and Uniform instances, as long as $T_0$ is sufficiently small. However, \TS is a weakly dominant strategy for the Needle-in-Haystack instance. We find that for a sufficiently small $T_0$, \DG yields more than half the market against \TS,  but achieves similar market share vs. \DG and \DEG. By our tie-breaking rule, (\DG,\DG) is the only pure-strategy equilibrium.

@@

\begin{finding}
\textit{When all firms deploy \DG, and start at the same time, welfare is weakly decreasing as the number of firms \gaedit{increases}}
\end{finding}
